Two subtle problems with over-representation analysis

### 2024-06-09 manuscript/manuscript.Rmd
ns

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running `enricher`.
After `enricher` using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* 1 values were appended to the p-values obtained by enricher,
followed by FDR correction in R, to obtain the corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be differentially
expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts` function of
the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the addition of some extra variation drawn from a
normal distribution with mean of 1 and standard deviation varying between 0.3 and 0.45, this
value selected randomly for each gene.
Five "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.

With these expression profiles, DESeq2 was used for differential analysis, and the results were
subjected to `enricher` which suffers from both problems, `fora`, an ORA function of the `fgsea`
bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a fast
implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes are detected as significant, then 200 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.

After enrichment analysis with these approaches, the significant sets up and down regulated sets
were selected and compared to the ground truth in order to calculate precision, recall and F1
score.

At each value of added variance, 500 replications were conducted, each with as set seed for
reproducibility.

@@

In the analysis of dataset 1, the background list contained 14,144 genes and the foreground
list consisted of 1,469 genes.
After ORA with KEGG sets, the background size reported by clusterProfiler was only 2,120.
With GO sets, the background was reported to be 12,365.
These numbers correspond to the number of genes in the background list that are annotated with
one or more functions in the respective gene set libraries.

We used a workaround to eliminate the background problem and compared the results to standard
analysis.
This was repeated for seven datasets and the nine gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).

>>>>

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running `enricher`.
After `enricher` using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
enricher, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 1.0, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes are detected as significant, then 200 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant sets up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected algorithm for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
